HEP Software Foundation Community White Paper Working Group -- Data Organization, Management and Access (DOMA)

### 2018-06-22 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex

The coming decade will bring HEP to the exabyte scale with the expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as the Square Kilometer Array (SKA) entering this
regime. In devising computing models for this era, many
factors have to be taken into account. In particular, the increasing
availability of very high-speed networks may reduce the need for
% * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
%
% > very high-speed networks
% What is "very high-speed" here? 100G? 1000G?
%
% ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
CPU and data co-location. Such networks may allow
for more extensive use of data access over the wide-area network (WAN) while providing fail-over capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation and analysis models, such as a potential move to event-based or
sub-event-based data streaming from the more traditional dataset-based or file-based
data access, will be particularly important for optimizing the
utilization of opportunistic computing cycles on HPC facilities,
commercial cloud resources, and campus clusters, and can potentially
resolve currently limiting factors such as job eviction.

Planned analysis of HL-LHC data and data analysis from other
% * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
%
% > large
% XENON1T has shown that the collaboration doesnt have to be "large" to require DOMA
%
% ^ <briedel@uchicago.edu> 2018-04-09T20:36:43.846Z.
international collaborations will need to adopt a distributed computing
model due to its scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

@@

The LHC experiments currently collectively provision and manage about an Exabyte of
storage of which approximately half is archival and half is
traditional disk storage. The annual storage requirements are expected
to jump by a factor of 10 for the HL-LHC. This increase is faster than
projected density gains in current storage technologies and thus will present major challenges. Storage
% * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
%
% > Moore's Law gains
% This is only a CPU. What about storage technology? Currently we are adding about 2 TB/year in disk capacity. It is unlikely will be at 100 TB/disk in 2023. We may have O(10) - O(100) TB SSDs  at a reasonable cost, but we can't be certain. What does the future of tape look like?
% changed to "density gains in current storage technologies." Tape scaling has quite a future but hard drives, less so
% ^.
will remain one of the visible cost drivers for HEP computing. The projected growth and cost of the computational resources needed to
analyze the data is expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
  the complexity of detector data and the computational algorithms
  needed to associate and disentangle the data. This is true of
  reconstruction methods used today at the LHC and throughout the HEP
  community and will continue to be the driving characteristic of
  reconstruction at the HL-LHC and for liquid argon TPC experiments. The
  CPU-bound nature of the algorithms and the ratio of event size to
  network bandwidth make the option of streaming individual events
  records (e.g., full collision events or beam spill triggers)
  across the WAN to compute elements a feasible
  strategy. The requirements of event streaming can be
  satisfied by a wide range of storage solutions.
\item \textbf{Simulation:} HEP event simulation strategies focus on
  transforming a small set of key input parameters into event and
  detector response information that is representative of what would be
  seen in the physical detector systems. Simulation processes are
  typically characterized by highly asymmetric IO. They typically have very small data ingest (typically just the key input
  parameters), while output data are similar in size and
  structure to real detector data. Due to their similar characteristics, output data from simulation processes share the requirements and associated challenges of detector data. The primary drivers that differentiate
  this type of HEP workflow from the other categorizations are the
  increased complexities that the simulation stages often require. These
  simulations often rely on auxiliary event generation mechanisms and
  can, in some HEP domains, require large external data sets
  representing interaction cross sections or detector response functions
  to be available to each instance of the event simulation. This type of
  common input and overlay data may present a challenge in relation to
  its scaling to the runtime environments of future HPC facilities and
  to the design of caching layers at computing site.
\item \textbf{Analysis:} Analysis level datasets and subsets are typically
  comprised of refined or reduced information that is most pertinent to
  the extraction of physics quantities. These analysis-focused event
  records can often be accepted or rejected through fast pre-filtered
  methods which do not require the reading and retrieval of the full
  event record. This fundamental difference in the way that the analysis
  data is consumed makes it more susceptible to storage bandwidth
  limitations and to IOPS transaction limits that underlying
  technologies can have. In particular, the internal structure of the
  event data can dramatically impact the efficiency of data retrieval
  and favor different access models and protocols while the actual analysis calculations being performed can
  dramatically affect the CPU-to-IO ratios. These factors make this
  categorization of data the most challenging to project to future
  storage needs and models.
\item \textbf{Replication:} Data replication, whether within a site or
  across multiple sites, is performed to enable efficient data access by
  improving data locality (the proximity of the data to compute
  resources) and exploiting the available data bandwidth between the
  storage and the compute locations. Data replication also serves to
  guard against data loss. Data replication places modest requirements
  on the IOPS that need to be supported, but drives
  the bandwidth requirements that are needed in enacting site-to-site
  data transfers between WAN endpoints (\emph{e.g.}, 10-100~Gbps). This
  requirement on the available WAN bandwidth provides a corresponding
  performance requirement on on the underlying storage systems to match
  network performance. Satisfying both requirements is necessary to
  provide efficient point-to-point data flow.
\end{enumerate}

@@

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (e.g. physics
object). Any such catalogs may have to be external to the primary data
store (e.g., relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
% * <briedel@uchicago.edu> 2018-04-05T02:04:26.292Z:
%
% > content-addressable
% This comes a little out of nowhere. Where is this discussed before?
%
% ^.
handle to have a fully descriptive location of a given event or other
object being sought.
% * <briedel@uchicago.edu> 2018-04-05T02:05:03.704Z:
%
% > or other object
% This is the first time "other object" is mentioned. What other object? TBranches? It would be better to call this sub-event data like done before.
%
% ^.

@@

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
% * <briedel@uchicago.edu> 2018-04-05T02:08:36.870Z:
%
% > Akamai.
% Google also offers one
%
% ^.
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance where a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products, minimizing data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role. NDNs allow the naming, at the
network level, of the data you are looking for rather than its location, which
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, Adaptive Replacement Cache-style caching) and may
% * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
%
% > server-less caching
% In-memory caching?
%
% ^.
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

@@

For more scalable analysis, another opportunity to evaluate is how much
work can be offloaded to a storage system, for example caching
uncompressed or reordered data for fast access. The idea can be extended
to virtual data and to query interfaces which would perform some of the
transformation logic currently executed on CPU workers. Interactive
querying of large datasets is an active field in the Big Data industry;
examples include Spark-SQL, Impala, Kudu, Hawq, Apache Drill, and Google
Dremel/BigQuery. A key question is about the usability of these
techniques in HEP and we need to assess if our data transformations are
not too complex for the SQL-based query languages used by these
products. We also need to take into account that the adoption of these
techniques, if they prove to be beneficial, would represent a disruptive
change which directly impacts the end user and therefore promoting
acceptance through intermediate solutions would be desirable. One such solution - ``Service X'' - has been proposed in \href{https://bit.ly/2Ix9NIL}{Organizing Data Lakes}. The service would start with the data in its persistent format, perform one or more operations on it  (\emph{e.g.} joins, query, decompression, filtering to sub-event level), and finally deliver a data stream to a client.

@@

HEP experiments should be prepared to leverage ``tactical storage''.
Such storage may be provisioned only when it becomes cost-effective
(e.g.,
% * <briedel@uchicago.edu> 2018-04-05T15:43:01.526Z:
%
% > most
% "more" rather than "most"?
%
% ^.
from a cloud provider) and have a data management and provisioning
system that can exploit such resources on short notice. Volatile data
% * <briedel@uchicago.edu> 2018-04-05T15:43:37.726Z:
%
% > Volatile data sources would impact many aspects of the system; catalogs, job brokering, monitoring/alerting, accounting, the applications themselves.
% Aren't we dealing with somewhat volatile storage in terms of xrootd caches today already? The file _might_ be there, but it isnt guaranteed.
%
% ^.
sources would impact many aspects of the system; catalogs, job
brokering, monitoring/alerting, accounting, the applications themselves.

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance trade-offs.
Currently, tape is extensively used to hold data that cannot economically be made available online. While the tape-stored data is still accessible,
it comes with a high latency penalty and thus limiting possible analysis. We suggest investigating either separate direct-access-based archives (e.g.
disk or optical) or new models that overlay online direct-access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.
% * <briedel@uchicago.edu> 2018-04-05T15:49:15.646Z:
%
% > stability.
% This paragraph begs the question of whether one needs to move to "direct-access-based archives" if the data is only accessed once or twice. The plenty of tape is not that great in that case.
%
% ^.

Cost reductions in maintenance and operation of the storage
infrastructure can be realized through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a mono-culture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialized systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavors that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

@@

\begin{itemize}
\item Evaluate row-based versus column-based
  data organization on the performance of each type of access used.
\item Investigate data storage and access solutions that support the use of map-reduce or
Spark-like analysis tools and their adaptability to HEP analysis needs.
\item Evaluate new compression schemes, just-in-time
  decompression schemes and mappings onto hardware architectures
  considering the flow of data from spinning disk to memory and
  application.
\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filtering through metadata
\end{itemize}

@@

-\emph{The first two actions should be completed by 2020 but evaluation of CDN/NDN is more long-term}

@@

\begin{itemize}
\item Understand data access patterns, including investigating frequency of access of data
\item Exploring ways to ease preliminary data filtering steps
\item Exploring commercially-popular data transfer methods (\emph{e.g,} HTTP)
\end{itemize}
\emph{The  proof-of-concept should be completed by 2020 and, if successful, implementation in
production will be done in the following years.}

>>>>

The coming decade will bring HEP to the exabyte scale with the expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as the Square Kilometre Array (SKA) entering this
regime. In devising computing models for this era, many
factors have to be taken into account. In particular, the increasing
availability of very high-speed networks may reduce the need for
CPU and data co-location. Such networks may allow
for more extensive use of data access over the wide-area network (WAN) while providing fail-over capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation and analysis models, such as a potential move to event-based or
sub-event-based data streaming from the more traditional dataset-based or file-based
data access, will be particularly important for optimizing the
utilization of opportunistic computing cycles on HPC facilities,
commercial cloud resources, and campus clusters, and can potentially
resolve currently limiting factors such as job eviction.

Planned analysis of HL-LHC data and data analysis from other
international collaborations will need to adopt a distributed computing
model due to their scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

@@

The LHC experiments currently collectively provision and manage about an Exabyte of
storage of which approximately half is archival and half is
traditional disk storage. The annual storage requirements are expected
to jump by a factor of 10 for the HL-LHC. This increase is faster than
projected density gains in current storage technologies and thus will present major challenges. Storage
will remain one of the visible cost drivers for HEP computing. The projected growth and cost of the computational resources needed to
analyze the data is expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
  the complexity of detector data and the computational algorithms
  needed to associate and disentangle the data. This is true of
  reconstruction methods used today at the LHC and throughout the HEP
  community and will continue to be the driving characteristic of
  reconstruction at the HL-LHC and for liquid argon TPC experiments. The
  CPU-bound nature of the algorithms and the ratio of event size to
  network bandwidth make the option of streaming individual events
  records (e.g., full collision events or beam spill triggers)
  across the WAN to compute elements a feasible
  strategy. The requirements of event streaming can be
  satisfied by a wide range of storage solutions.
\item \textbf{Simulation:} HEP event simulation strategies focus on
  transforming a small set of key input parameters into event and
  detector response information that is representative of what would be
  seen in the physical detector systems. Simulation processes are
  typically characterized by highly asymmetric IO. They typically have very small data ingest (typically just the key input
  parameters), while output data are similar in size and
  structure to real detector data. Due to their similar characteristics, output data from simulation processes share the requirements and associated challenges of detector data. The primary drivers that differentiate
  this type of HEP workflow from the other categorizations are the
  increased complexities that the simulation stages often require. These
  simulations often rely on auxiliary event generation mechanisms and
  can, in some HEP domains, require large external data sets
  representing interaction cross sections or detector response functions
  to be available to each instance of the event simulation. This type of
  common input and overlay data may present a challenge in relation to
  its scaling to the runtime environments of future HPC facilities and
  to the design of caching layers at computing site.
\item \textbf{Analysis:} Analysis level datasets and subsets are typically
  comprised of refined or reduced information that is most pertinent to
  the extraction of physics quantities. These analysis-focused event
  records can often be accepted or rejected through fast pre-filtered
  methods which do not require the reading and retrieval of the full
  event record. This fundamental difference in the way that the analysis
  data is consumed makes it more susceptible to storage bandwidth
  limitations and to IOPS transaction limits that underlying
  technologies can have. In particular, the internal structure of the
  event data can dramatically impact the efficiency of data retrieval
  and favor different access models and protocols while the actual analysis calculations being performed can
  dramatically affect the CPU-to-IO ratios. These factors make this
  categorization of data the most challenging to project to future
  storage needs and models.
\item \textbf{Replication:} Data replication, whether within a site or
  across multiple sites, is performed to enable efficient data access by
  improving data locality (the proximity of the data to compute
  resources) and exploiting the available data bandwidth between the
  storage and the compute locations. Data replication also serves to
  guard against data loss. Data replication places modest requirements
  on the IOPS that need to be supported, but drives
  the bandwidth requirements that are needed in enacting site-to-site
  data transfers between WAN endpoints (e.g., 10-100~Gbps). This
  requirement on the available WAN bandwidth provides a corresponding
  performance requirement on on the underlying storage systems to match
  network performance. Satisfying both requirements is necessary to
  provide efficient point-to-point data flow.
\end{enumerate}

@@

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (e.g. physics
object). Any such catalogs may have to be external to the primary data
store (e.g., relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
handle to have a fully descriptive location of a given event or other
object being sought.

@@

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance where a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products, minimizing data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role. NDNs allow the naming, at the
network level, of the data you are looking for rather than its location, which
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, ARC-style caching) and may
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

@@

For more scalable analysis, another opportunity to evaluate is how much
work can be offloaded to a storage system, for example caching
uncompressed or reordered data for fast access. The idea can be extended
to virtual data and to query interfaces which would perform some of the
transformation logic currently executed on CPU workers. Interactive
querying of large datasets is an active field in the Big Data industry;
examples include Spark-SQL, Impala, Kudu, Hawq, Apache Drill, and Google
Dremel/BigQuery. A key question is about the usability of these
techniques in HEP and we need to assess if our data transformations are
not too complex for the SQL-based query languages used by these
products. We also need to take into account that the adoption of these
techniques, if they prove to be beneficial, would represent a disruptive
change which directly impacts the end user and therefore promoting
acceptance through intermediate solutions would be desirable. One such solution - ``Service X'' - has been proposed in \href{https://bit.ly/2Ix9NIL}{Organizing Data Lakes}. The service would start with the data in its persistent format, perform one or more operations on it  (e.g. joins, query, decompression, filtering to sub-event level), and finally deliver a data stream to a client.

@@

HEP experiments should be prepared to leverage ``tactical storage''.
Such storage may be provisioned only when it becomes cost-effective
(e.g.,
from a cloud provider) and have a data management and provisioning
system that can exploit such resources on short notice. Volatile data
sources would impact many aspects of the system; catalogs, job
brokering, monitoring/alerting, accounting, the applications themselves.

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance trade-offs.
Currently, tape is extensively used to hold data that cannot economically be made available online. While the tape-stored data is still accessible,
it comes with a high latency penalty and thus limiting possible analysis. We suggest investigating either separate direct-access-based archives (e.g.
disk or optical) or new models that overlay online direct-access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.

Cost reductions in maintenance and operation of the storage
infrastructure can be realized through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a monoculture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialized systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavors that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

@@

\begin{itemize}
\item Evaluate row-based versus column-based
  data organization on the performance of each type of access used.
\item Investigate data storage and access solutions that support the use of map-reduce or
Spark-like analysis tools and their adaptability to HEP analysis needs.
\item Evaluate new compression schemes, just-in-time
  decompression schemes and mappings onto hardware architectures
  considering the flow of data from spinning disk to memory and
  application.
\item Evaluate possible alternative storage formats to the ROOT format, especially storage formats optimized for storage efficiency and/or  data filtering through metadata
\end{itemize}

@@

\emph{The first two actions should be completed by 2020 but evaluation of CDN/NDN is more long-term}

@@

\begin{itemize}
\item Understand data access patterns, including investigating frequency of access of data
\item Exploring ways to ease preliminary data filtering steps
\item Exploring commercially-popular data transfer methods (e.g. HTTP)
\end{itemize}
\emph{The  proof-of-concept should be completed by 2020 and, if successful, implementation in
production will be done in the following years.}
